Analysis of gene regulatory properties underlying trait pleiotropy

### 2022-10-15 presentation_gold2022_paris/presentation_gold2022_paris.tex
ENH presentation gold meeting

\begin{itemize}
\item Genetic loci with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Limitations of GWAS}

\begin{figure}[!]
\includegraphics[width=0.25\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/41586_2017_Article_BFnature24284_Fig1_HTML.jpg}
\caption{GWAS for breast cancer}
\end{figure}

\begin{itemize}
\item Correlation between variants, ie. LD
\item Which cell types are causal to the disease?
\item Over 90\% GWAS variants fall in non-coding regions
\end{itemize}
%
\vfill
%
These points can be partially addressed by looking at eQTLs in different tissues

\let\thefootnote\relax\footnotetext{doi:10.3389/fgene.2020.00424}
\end{frame}

@@

\section{Objectives} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Materials and Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Data and Algorithm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Which are the most pleiotropic variants?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants have more egenes?}

\begin{figure}[!tbp]
\centering
%
\begin{subfigure}[]{.32\textwidth}
\textbf{a}
\\
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_etissue_y_egene.py/plt.png}
\end{subfigure}
%
\begin{subfigure}[]{.32\textwidth}
\textbf{b}
\\
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_egene_y_etissue.py/plt.png}
\end{subfigure}
%
\begin{subfigure}[]{.32\textwidth}
\textbf{c}
\\
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_egene_etissue_y_gwas.py/plt.png}
\end{subfigure}
%
\caption{Count of eGenes per variant-tissue (\textbf{a}), eTissues per variant-eGene (\textbf{b}) and GWAS categories per variant-eGene-eTissue (\textbf{c}).} \label{fig:gwas_egene_etisue_per_variant}
%
\end{figure}

@@

\begin{itemize}
\item A  colocalisation pipeline to link EBI Catalogue eQTLs and IEA OpenGWAS signals
\item Several colocalisation signals between immune cell eQTLs and cancer loci
\item Preliminary analyses of these colocalised signals
\end{itemize}

@@

\begin{itemize}
\item Use H1 as background dataset: H1=Causal SNP in cancer w/out eQTL
\item Statistics of cell type enrichments
\item Causal SNP characterization: Motif analyses
\item Other applications:
\begin{itemize}
\item Pleiotropy of eQTLs: eQTLs affecting two genes or same gene in two cell-types
\item Causality between other cell type eQTLs and diseases: Immunity and Covid-19
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item L\'eopoldine Lecerf (M1)
\item P Paul, P Rihet, M Michel
\item S Marquet, S Spicuglia
\item Funding:
\begin{itemize}
\item Institut Cancer et Immunologie (ICI)
\item Agence nationale de la recherche (ANR)
\end{itemize}
\end{itemize}

>>>>

\begin{itemize}
\item Genetic loci with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}
%
\vfill
%
Limitations
%
\vfill
%
\begin{itemize}
\item Correlation between variants, ie. LD
\item Which cell types are causal to the disease?
\item Over 90\% GWAS variants fall in non-coding regions
\end{itemize}
%
\vfill
%
These limitations can be partially addressed by looking at eQTLs in different tissues

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Data and Algorithms}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{How are pleiotropic variants distributed?}

\begin{columns}
\begin{column}{0.32\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltsctr_x_per_rsid_y_gwas.py/count_per_rsid_chr19_start48668049_end48756272_categories8.png}
     \end{center}
\end{column}
\begin{column}{0.32\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltsctr_x_per_rsid_y_egene.py/count_per_rsid_chr19_start48668049_end48756272_categories8.png}
     \end{center}
\end{column}
\begin{column}{0.32\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltsctr_x_per_rsid_y_egene.py/count_per_rsid_chr19_start48668049_end48756272_categories8.png}
     \end{center}
\end{column}
\end{columns}
%
\vfill
%
Neighboring variants show similar pleiotropy.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Where are the pleiotropic variants?}

    \begin{center}
\includegraphics[width=0.8\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_vep_consequence.py/vep.png}
     \end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more egenes?}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=0.7\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/model_pleio_egenes.png}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}  %%<--- here
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_etissue_y_egene.py/plt.png}
     \end{center}
\end{column}
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more etissues?}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=0.7\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/model_pleio_etissues.png}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}  %%<--- here
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_egene_y_etissue.py/plt.png}
     \end{center}
\end{column}
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more traits (after fixing egene-etissue)?}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=0.7\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/model_pleio_gwas.png}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}  %%<--- here
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_egene_etissue_y_gwas.py/plt.png}
     \end{center}
\end{column}
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants bind more transcription factors?}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbox_x_per_rsid_y_remapnr.py/bxplt_remaptf_per_rsid_flank_10.png}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}  %%<--- here
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_x_per_variant_pleiotropy_y_remapcrm.py/remapcrm_flank10.png}
     \end{center}
\end{column}
\end{columns}

@@

Characteristics of pleiotropic variants
%
\begin{itemize}
\item Variant pleiotropy is correlated in regions
\item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
\item Enrichment of transcription factor and CRM binding
\item Increased number of associated egenes and etissues
\end{itemize}
%
\vfill
%
Other outcomes of this study
%
\begin{itemize}
\item A large analysis of GWAS/eQTL colocalisations
\item A list of pleiotropic variants and regions
\item Analysis of regulatory properties of pleiotropic variants
\end{itemize}

@@

\begin{itemize}
\item A web portail to share data
\item Causality of egenes on gwas (Looking for experts ;-) )
\end{itemize}

@@

\begin{itemize}
\item L\'eopoldine Lecerf (M1)
\item P Paul, P Rihet, M Michel, S Marquet, S Spicuglia
\end{itemize}
%
\vfill
%
Funding
%
\begin{itemize}
\item Institut Cancer et Immunologie (ICI)
\item Agence nationale de la recherche (ANR)
\item GOLD
\end{itemize}
